---
title: "data-wrangling"
output: pdf_document
---
## Responsibilities:
Skye — data import, shapefile, incarceration trends, time series
Diego — incarceration by state, rates by race, ratio
Alexander — incarceration trends map

```{r setup, include = FALSE}
# load packages
library(readxl)
library(mdsr)
library(tidyverse)
library(dplyr)
library(knitr)
library(ggplot2)
library(janitor)
library(vroom) # fast reading/importing data
library(sf) # spacial data
library(tigris) # geojoin
library(leaflet) # interactive maps
library(htmlwidgets) # interactive map labels
library(readr)
library(rmapshaper)

# set code chunk defaults
knitr::opts_chunk$set(tidy = F, # display code as typed
                      size = "small", # slightly smaller code font
                      message = FALSE,
                      warning = FALSE,
                      comment = "\t") 
# set black & white default plot theme 
# theme_set(theme_classic()) # uncomment this line if desired
# improve digit and NA display 
options(scipen = 1, knitr.kable.NA = '')
```

## Data Import
```{r, eval = TRUE}
# Incarceration Trends
incarceration_trends <- vroom("incarceration_trends.csv")
incarceration_rates <- read_csv("prison_policy_initiative_incarceration_by_race.csv")
incarceration_populations <- read_csv("bjs_total_correctional_population_counts_by_status_TM 5_26_21.csv")
incarceration_rates_race <- read_csv("prison_policy_initiative_incarceration_by_race.csv") # data from 2010
rates_by_race <- read_csv("sentencing_project_rates_by_race.csv")
ratio_bw <- read_csv("ratio_bw.csv")
```

## Shapefile Wrangling
```{r, eval = TRUE}
# read in shapefile
unzip(zipfile = "geography resource/cb_2018_us_county_500k.zip")
county_shapefile <- read_sf("cb_2018_us_county_500k.shp") %>%
  janitor::clean_names() 

# take out statefp and countyfp and rename geoid column into fips
county_shapefile <- county_shapefile %>%
  select(- statefp, - countyfp) %>%
  mutate(fips = as.numeric(geoid)) %>%
  ms_simplify() #this function from rmapshaper significantly speeds up the map by making the county polygons less complicated

write_rds(county_shapefile, "county_shapefile_wrangled.rds")

# save the wrangled shapefile
# st_write(county_shapefile, "county_shapefile_wrangled.shp") # only run once, so the file does
#not get overwritten each time
```

## Incarceration Trends Wrangling
```{r, eval = TRUE}
# add a space between words in county names
incarceration_trends <- incarceration_trends %>%
  separate(county_name, c("county", NA), sep = " ")

write_csv(incarceration_trends, "incarceration_trends_wrangled.csv")
```

## Incarceration by State Wrangling
```{r, eval = TRUE}
#rename necessary columns
incarceration_rates_race <- incarceration_rates_race %>% 
  rename(ID = "...3",
         rate_per_100 = "...26") %>% 
  #delete United States row and remove states and territories not in Continental United States
  filter(ID != "United States", 
         ID != "Puerto Rico", 
         ID != "Geography", 
         ID != "Hawaii", 
         ID != "Alaska") %>%
#Make state names lowercase to match geom dataset and change rate column to numeric
  mutate(ID = tolower(ID),
         #Take out commas from rates column
         rate_per_100 = gsub(",","", as.character(rate_per_100)),
         rate_per_100 = as.numeric(rate_per_100))
```

#Population Time-Series wrangling
```{r, eval = TRUE}
incarceration_populations <- incarceration_populations %>% 
  # rename the columns that are numbers
  rename(total_incarcerated = "...7",
         year = "Bureau of Justice Statistics",
         total_supervised = "...2",
         local_jail = "...8",
         state_prison = "...10",
         federal_prison = "...11") %>%
  # take out empty rows at the top
  filter(total_incarcerated != "Total/c", !row_number() %in% c(1)) %>%
  separate(year, c("year", NA), sep = "/d") %>%
  # take out "," in the population counts and convert them to doubles
  mutate(
    total_incarcerated = as.numeric(gsub(",","", as.character(total_incarcerated))),
    total_supervised = as.numeric(gsub(",","", as.character(total_supervised))),
    local_jail = as.numeric(gsub(",","", as.character(local_jail))),
    state_prison = as.numeric(gsub(",","", as.character(state_prison))),
    federal_prison = as.numeric(gsub(",","", as.character(federal_prison))),
    year = as.numeric(year))
```

#Incarceration Rates by Race wrangling
```{r, eval = TRUE}
# define a function to capitalize each word of state names
simpleCap <- function(x) {
  s <- strsplit(x, " ")[[1]]
  paste(toupper(substring(s, 1, 1)), substring(s, 2),
        sep = "", collapse = " ")
}

rates_by_race <- rates_by_race %>% 
  # change variable type of the rates of each race
  mutate(Black = as.numeric(Black),
         Latino = as.numeric(Latino),
         White = as.numeric(White),
         # change the state names using the defined function
         State = sapply(State, simpleCap)) %>%
  arrange(State) %>%
  # pivot longer for graphing
  pivot_longer(!State, names_to = "race", values_to = "rate")

write_csv(rates_by_race, "rates_by_race_wrangled.csv")
```

#Incarceration Rates Wrangling and File Splitting for Map
```{r,eval = TRUE}
# only keep necessary columns
incarceration_trends_map <- incarceration_trends %>% 
  select (yfips, year, fips, state, county, total_pop, total_prison_pop_rate, female_prison_pop_rate, male_prison_pop_rate, aapi_prison_pop_rate, black_prison_pop_rate, latinx_prison_pop_rate, native_prison_pop_rate, white_prison_pop_rate, total_jail_pop_rate, female_jail_pop_rate, male_jail_pop_rate, aapi_jail_pop_rate, black_jail_pop_rate, latinx_jail_pop_rate, native_jail_pop_rate, white_jail_pop_rate)

incarceration_trends_map <- incarceration_trends_map %>% 
  rowwise() %>%
  # add up prison and jail rates to create new sum columns
  mutate(
    total_jail_prison_pop_rate = sum(total_prison_pop_rate, total_jail_pop_rate, na.rm = TRUE),
    female_jail_prison_pop_rate = sum(female_prison_pop_rate, female_jail_pop_rate,na.rm = TRUE),
    male_jail_prison_pop_rate = sum(male_prison_pop_rate, male_jail_pop_rate, na.rm = TRUE),
    aapi_jail_prison_pop_rate = sum(aapi_prison_pop_rate, aapi_jail_pop_rate, na.rm = TRUE),
    black_jail_prison_pop_rate = sum(black_prison_pop_rate, black_jail_pop_rate, na.rm = TRUE),
    latinx_jail_prison_pop_rate = sum(latinx_prison_pop_rate, latinx_jail_pop_rate, na.rm = TRUE),
    native_jail_prison_pop_rate = sum(native_prison_pop_rate, native_jail_pop_rate, na.rm = TRUE),
    white_jail_prison_pop_rate = sum(white_prison_pop_rate, white_jail_pop_rate, na.rm = TRUE)
  )

write_csv(incarceration_trends_map, "map_data/all_years_incarceration_trends_map.csv")

# group the entries by year
incarceration_trends_map <- incarceration_trends_map %>% 
  group_by(year)

group_walk(incarceration_trends_map, ~ write_csv(.x, paste0("map_data/",.y$year, "_incarceration_trends_map.csv")))

```

#Ratio Wrangling
```{r, eval = TRUE}
# rename certain variables
ratio_bw <- ratio_bw %>% 
  rename(ID = State,
        ratio = "Ratio (B/W)") %>%
  # change variable types
  mutate(ID = tolower(ID),
        ratio = as.numeric(ratio))
```

